---
title: "Module 5"
subtitle: "Basic data visualization in RStudio"
project:
  type: website
  output-dir: docs
---

```{r setup, echo=FALSE, include=F}

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

setwd("C:/Users/Work/OneDrive - Umeå universitet/NWU_2025_workshop")

library(dplyr)
library(data.table)
library(tidyverse)
library(ggplot2)

enterobase_assemblystats<-fread("C:/Users/Work/OneDrive - Umeå universitet/NWU_2025_workshop/20251020_enterobase_minnesota_assemblystats.tsv",header=T,sep="\t")

enterobase_mlst<-fread("C:/Users/Work/OneDrive - Umeå universitet/NWU_2025_workshop/20251020_enterobase_minnesota_mlst.tsv",header=T,sep="\t")

```

R is very commonly used in scientific research for statistical analysis and data visualization for many different types of data modalities.

R is a programming language, whereas Rstudio is an **Integrated development environment** or IDE.

The goal of an IDE is to provide a single environment for writing and running code, troubleshooting/debugging, package management as well as rendering visualizations.

In this workshop, we will primarily be using R/Rstudio for data visualization. I will quickly get to handling real tabular data and creating plots while explaining some basic concepts along the way. However, if you would like a comprehensive tutorial on the basics of R from scratch, please check out the [Software Carpentry R tutorial](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html) and/or [Swirl](https://swirlstats.com/)

Your RStudio will look something like this:

![[RStudio Panes](https://docs.posit.co/ide/user/ide/get-started/)](https://docs.posit.co/ide/user/ide/get-started/images/rstudio-panes-labeled.jpeg)

You can write your code in the **Source** pane and press `ctrl + enter` to run the selected line(s). This is your scripting environment where you can save everything you are typing.

Alternatively, you can also write code in the **Console** pane and press `enter` to run as you were doing for the bash command line. **The code in the console is not saved.**

Any datasets you load or variables/functions you set will be visible in the **Environments** pane.

Any plots/documents that you render will be visible in the **Output** pane.

### Setting up your R environment

New packages can be installed using `install.packages("package_name")`

Not all installed R packages are available by default. You can load packages using the `library` function.

The R working directory can be seen using the `getwd()` command

You can change the working directory using `setwd("/path/to/dir")`

``` r

install.packages(c("data.table","dplyr","tidyverse","ggplot2"))

library(dplyr)
library(tidyverse)
library(ggplot2)

getwd()

setwd("NWU_2025_workshop_data/test_datasets/results")
```

### Download an example table

We learned about Enterobase in Module 3. On your browser, navigate to the Enterobase *Salmonella* page, and search for the serovar 'Minnesota'. Download the assemblystats table and the Achtman 7 Gene MLST table. 

### Reading your data

We are using the **function** `fread` which is part of the `data.table` package to read a table.

To use **functions** in R, you call the name of the function, and enclose the different values the function will accept within parenthesis.

To access the help page for any function, type `?` followed by the name of the function

``` r
?fread
```

`fread` is a function to quickly read `.tsv` or `.csv` files

#### Lets read in our Enterobase assemblystats table

``` r
enterobase_assemblystats<-fread(file="/path/to/20251020_enterobase_minnesota_assemblystats.tsv",header=T,sep="\t")
```

-   `fread(...)` : Calling the `fread` function
-   `file="..."` : Providing the path for the file to be read
-   `header=T`: Stating that the file has a header (which is the first line in the `csv` or `tsv` that has all the column names). `T` is short for `TRUE`.
-   `sep="\t"`: Stating the delimiter in the file, since our file is tab separated, we set the delimiter as `\t`.
-   `enterobase_assemblystats<-`: We are saving the read table to an object called `enterobase_assemblystats` . Once it is saved, it should appear in your **Environment** pane. You can click on it here and it will open a new tab in your **Source** pane where you can browse through the table.

### Making simple plots using ggplot2

Ggplot is the most common R tool used for creating publication quality plots.

ggplot follows a consistent grammar (The **gg** in ggplot means **g**rammar of **g**raphics) making it very easy to build upon.

Apart from the default ggplot options, there are several additional packages using the ggplot grammar such as [ggtree](https://guangchuangyu.github.io/software/ggtree/), [ggside](https://github.com/jtlandis/ggside), [ggsignif](https://const-ae.github.io/ggsignif/) and many others that are all very helpful in scientific data visualization. ggplot can make (*almost*) any kind of plot you can think of!

#### Making a histogram using ggplot2

Lets make a simple histogram showing the distribution of the number of contigs in our dataset

```{r}
ggplot(data = enterobase_assemblystats,aes(x=`Contig Number(>=200 bp)`))+
  geom_histogram(binwidth = 5)
```


-   `ggplot()` : Calling the ggplot function
-   `data = enterobase_assemblystats`: The data that we want to use, here it is `enterobase_assemblystats`, which is the enterobase table that we read
-   `aes()` : These are the mapping aesthetics (i.e which column of the input dataset we want our x and y axes to be)
    -   `` x=`Contig Number(>=200 bp)` `` : The name of the column we want on the `x` axis.
    -   Unfortunately, the table has column names with spaces and special characters like `(>=)` which can have different special meaning. So we have to take some extra steps to make sure the column names are read properly.
    -   Here we enclose the column name in \`\` - these are called "backticks".
    -   Now, the column name will be read literally and the special characters will not have special meanings.
    -   Normally you could simply say `x=column_name` if your column names don't have any special characters
-   `+ geom_histogram(binwidth = 5)` - Adding a new layer
    -   In ggplot, the `+` symbol is used to signify the addition of a new layer. We used `ggplot()` to tell it what data to use, and now we use `geom_` to tell it how to visually represent the data.
    -   `geom_histogram` is the histogram function in ggplot.
    -   `binwidth = 5` is a value for the width of the histogram "bars" that `geom_histogram` accepts. This is the range of values that are represented by one bar in the histogram. Increase or decrease this number to see how it changes the plot!

#### Making a scatterplot using ggplot2

Make a scatter plot (`geom_point`) with the number of contigs on the x axis and the N50 value in the y axis

<details> 

<summary> Click to reveal answer! </summary>

```{r}
ggplot(data = enterobase_assemblystats,aes(x=`Contig Number(>=200 bp)`,y=N50))+
  geom_point()
```

-   `ggplot()` : Calling the ggplot function
-   `data = enterobase_assemblystats`: The data that we want to use, here it is `enterobase_assemblystats`, which is the QUAST output table that we read
-   `aes()` : These are the mapping aesthetics (i.e which column of the input dataset we want our x and y axes to be)
    -   `` x=`Contig Number(>=200 bp), y=N50` `` : The name of the columns we want on the `x` and `y` axes.
-   `+ geom_point()` - Adding the scatterplot function in ggplot

</details>

-   Looks like there are two complete genomes in the table with very high N50 values that's throwing off the scale! 

-   We can make the plot with only samples having N50 values less than 4 million (<4e+06).

-   This means we have to subset the `enterobase_assemblystats` table, only keeping rows where the `N50` column has a value less than 4 million

-   We can do this like so:

```{.r}
enterobase_assemblystats[enterobase_assemblystats$N50 < 4e+06,]
```
  
  - `enterobase_assemblystats$N50` : The `$` sign is used to call a specific column within the dataframe. Here we are calling the "N50" column. 
  - `< 4e+06` means 'less than 4 million'. This is our **condition**
  - `enterobase_assemblystats$N50 < 4e+06` : Type this in the console and look at the output! You will see a list (in R, this is called a **vector**)  of 'TRUE' and/or 'FALSE' values. It is telling us for which rows the above condition (i.e, N50 < 4 million) is TRUE (or FALSE)
  - But we don't just want to know for which rows the condition is TRUE for, we want to filter the dataframe such that we get all the rows where the **condition is true**.
  - `enterobase_assemblystats[enterobase_assemblystats$N50 < 4e+06,]` : This is how you **filter** a dataframe in R.
    - The format is `dataframe[rows,columns]`
    - The rows we want and the columns we want are split by a **comma**
    - Instead of `rows` - we have our **condition** which is all the rows where N50 < 4e+06.
    - We are not filtering by columns, we want all columns so we leave it blank. That's why there is nothing after the **comma**
    
:::{.callout-tip collapse=true title="BONUS: Filter by columns too!"}
        `enterobase_assemblystats[enterobase_assemblystats$N50 < 4e+06,c(1,4,7)]` will print only the first, fourth and seventh column of the dataframe. After the **comma** we specify the columns we want, enclosed in `c()`. 
        You can run `colnames(dataframe)` to see the names of all the columns (and the column number). Run `colnames(enterobase_assemblystats)` on the console and find the number for the N50 column!
:::

We can thereby alter the `ggplot` command this:
Note that it is identical to the previous `ggplot` command, we have only changed what we give as input to `data`.

```{r}
ggplot(data = enterobase_assemblystats[enterobase_assemblystats$N50<4000000,],aes(x=`Contig Number(>=200 bp)`,y=N50))+
  geom_point()
```

Try playing around with different x and y axes, and also different filters. You can also add a `color` / `fill` to `aes` to colour your datapoints/bars based on a specific factor. 

#### EXERCISE: Remake the N50 vs Number of contigs scatter plot colouring the points by 'Country' !

<details>
<summary> Click here to see the answer! </summary>

```{r}
ggplot(data = enterobase_assemblystats[enterobase_assemblystats$N50<4000000,],aes(x=`Contig Number(>=200 bp)`,y=N50,color=Country))+
  geom_point()
```

</details>


#### EXERCISE: Make a boxplot showing the distribution of N50 values per country using `geom_boxplot`

<details>
<summary> Click here to see the answer! </summary>

```{r}
ggplot(data = enterobase_assemblystats[enterobase_assemblystats$N50<4000000,],aes(x=N50,y=Country))+
  geom_boxplot()
```
</details>

### Data cleaning/wrangling

We saw how to subset/filter a dataframe. Lets look at some more examples

#### Get rid of NAs

You will see that your enterobase table has many missing or "NA" values. We can select only specific rows that do not have missing values. For eg: Lets select only the "Uberstrain","Source Niche", "Collection Year", "Country", and "Sample ID" columns and remove all rows with 'NA' such that only rows where all four columns have values will be present

  - First select only the columns we want. Use `colnames(enterobase_assemblystats)` to get the column numbers for our desired columns (alternatively you can also specify the names in quotes separated by commas within `c()` ).
  - Write the subsetted dataframe to a new dataframe using `<-`

    ```{r}
    enterobase_assemblystats_filtered<-enterobase_assemblystats[,c(1,5,8,13,29)]
    ```
  - Now remove all "NA" values from the new dataframe.
    ```{r}
    enterobase_assemblystats_filtered<-enterobase_assemblystats_filtered[complete.cases(enterobase_assemblystats_filtered)]
    ```


#### Cleaning non-standard column values ("unknown" "Unknown" "ND")

You can see in the 'Source Niche' column in the `filtered` dataframe that there are still some blank values that did not get marked appropriately as 'NA'. There are also some columns marked 'ND' which is effectively the same as a missing value. These types of errors are common with public data which is why you must always double check after filtering. 

**How do you check if all the values in a column are OK?**

An easy trick is to use the `table` command, which produces frequency tables. That is, it prints out every value in a column along with the number of times that value has occurred.

```{r}
table(enterobase_assemblystats_filtered$`Source Niche`)
```

Here we can see a blank value has occurred 62 times and "ND" has occurred 20 times in the filtered set. Lets remove all rows where the "Source Niche" column has value = "" (blank) or = "ND"

```{r}
enterobase_assemblystats_filtered<-enterobase_assemblystats_filtered[!enterobase_assemblystats_filtered$`Source Niche` %in% c("","ND"),]
```

  - Here, we followed the same format for filtering the dataframe as we did with the N50 values: `dataframe[rows,columns]`
  - For the `rows`, we entered `` !enterobase_assemblystats_filtered$`Source Niche` %in% c("","ND") ``
    - `!` means **NOT** , so by starting the **condition** with `!`, we are saying "For rows where the following condition is **NOT** TRUE
    - `` enterobase_assemblystats_filtered$`Source Niche` `` : Refers to the "Source Niche" column, enclosed in backticks because of the space character
    - `%in% c("","ND")` : means if the value is **in** the list `c("","ND")` - which are the values we are looking for ("" is blank, and "ND" is ND)
    - So the full command means, Get all rows from the dataframe "enterobase_assemblystats_filtered"  IF the value in the "Source Niche" column is NOT "" or "ND"

:::{.callout-tip collapse=true title="Filter your dataframe as you read the table"}
Alternatively, if you know what you are looking for, you can appropriately specify what values should be counted as `NA` right when you read your dataframe using the `na.strings=` option.

```{.r}
enterobase_assemblystats<-fread("/home/vishnu/UmeaU_backup/NWU_2025_workshop/NWU_2025_workshop_data/test_datasets/20251020_enterobase_minnesota_assemblystats.tsv",na.strings = c("ND","","Unkown","unknown"))
```

:::

#### Merging data tables

You can merge two different data frames by a common column using `merge`

Read in the enterobase 7 Gene MLST table you downloaded

The advantage with `fread` is that you can select exactly which columns to read in right at the start so that the whole table isn't read. In our case, we only want column 1 and 34, so we specify that as `select = c(1,34)`.

```{.r}
enterobase_mlst<-fread(file="/path/to/20251020_enterobase_minnesota_mlst.tsv",header=T,sep="\t",select = c(1,34))
```

Lets merge the MLST table with our filtered assemblystats table using the `merge` command

```{.r}
enterobase_assemblystats_mlst_merged<-merge(enterobase_assemblystats_filtered,enterobase_mlst,by="Uberstrain")
```

  - The `merge` function is used like so: `merge(dataframe_1,dataframe_2,by="column_name")`
  - As long as the column name specified is the same for both our dataframes (in our case both columns are named "Uberstrain"), it will merge, keeping only values common in both dataframes. 

#### Adding a column by condition

You can add a new column based on a specific condition using the `mutate` function. For example lets say we want a column called 'is_human' which will have the values "yes" if "Source Niche" equals "Human" and "no" if not. 

```{.r}
enterobase_assemblystats_mlst_merged<-mutate(enterobase_assemblystats_mlst_merged, is_human = if_else(`Source Niche` == "Human", "yes", "no"))
```
  - We used the `mutate` function like so: `mutate(dataframe, <name for new column> = <value for new column>)`
  - We named the new column `is_human` , and used an `if_else` condition for the value
  - `` if_else(`Source Niche` == "Human", "yes", "no") `` : `if_else(<CONDITION>,<What to do if TRUE>,<What to do if FALSE>)` 
    - Here the CONDITION is IF "Source Niche" equals (`==`) "Human". If true, then the value for the new column `is_human` is "yes", else, the value is "no"

#### Group by and summarize

If you want the mean of a column, say N50, you can simply do:
```{r}
mean(enterobase_assemblystats$N50, na.rm = TRUE)
```

The `na.rm` option is to remove the `NA` values.

But what if you want the mean N50 value for each country? That is where the function `group_by` is used
```{r}
enterobase_assemblystats %>% group_by(Country) %>%  summarise(mean_N50_per_country = mean(N50, na.rm = TRUE))
```

- `%>%` is similar to the `|` (pipe) character in bash.
- `enterobase_assemblystats %>% group_by()` pipes the enterobase_assemblystats table to the group_by function
- `group_by(Country) %>%` : we want to group the different rows from the input by the "Country" column. This creates as many "groups" as there are countries. When we pipe this out, each "group" will be treated separately.
- `summarise(mean_N50_per_country = mean(N50, na.rm = TRUE))` : 
    - We piped the "groups" to the `summarise` function, which will now provide a "summary" for each group. 
    - We are asking it to create a new column `mean_N50_per_country`
    - `mean(N50, na.rm = TRUE)` is the same `mean` function we used above. We are using the `mean` function to get the mean of N50 values while removing any missing values using `na.rm=TRUE`
- This provides an output of "Grouped" means, each group being a country.    

## Sources and further reading

  - Enterobase: <https://enterobase.warwick.ac.uk/>
  - GGplot2 Essentials from STHDA: <https://www.sthda.com/english/wiki/ggplot2-essentials>
  - R Graph Gallery: <https://r-graph-gallery.com/ggplot2-package.html>
  - Software Carpentry R course: <https://swcarpentry.github.io/r-novice-gapminder/index.html>
